Methods that remove batch effects while retaining group differences may lead to exaggerated confidence in downstream analyses

### 2014-06-18 tex/manuscript.tex
Updated manuscript (main tex and pdf files)
Have merged (manually) the two different manuscript versions, and placed
as manuscript.tex, adding PDF and placing the figure files in the Fig
subfolder (renaming two of the files). Note that I do not now do the
LaTeX typesetting inside the Git folder, so the auxhiliary files are not
included including the tex-aux folder.

%% NOTES TO SELF:
% Cite using:
%    \citet{...}  in text
%    \citealp{...}   within parantheses
%
% Temporary notifications and reminders:
\newcommand\NB[1]{{\color{red}#1}}
\newcommand\NOTE[1]{{\color{red}[#1]}}
\newcommand\CITE[1]{{\color{magenta}[#1]}}

@@

@@

% List of authors, with corresponding author marked by asterisk
\author{VEGARD NYGAARD, EINAR ANDREAS RØDLAND, EIVIND HOVIG$^\ast$\\[4pt]
% Author addresses
\textit{Department of Tumor Biology,
Institute for Cancer Research,
Oslo University Hospital HF - Radiumhospitalet,
Montebello,
0310 Oslo,
Norway}
\\[2pt]
% E-mail address for correspondence
{ehovig@radium.uio.no}}

@@

The standard way to handle an extraneous variable is to include it in the statistical model employed in the inquiry. However, many analysis tools for high throughput data do not cater for this option, and when available it could still be outside the competence of the investigator. Therefore, an alternative two step procedure has emerged. First the batch effects are estimated and removed, creating a ``batch effect free'' data set. In the next step, the statistical analyses are performed on the adjusted data without further consideration of batch effects. This appealing compartmentalization is also convenient for practical purposes, for example when data-processing and statistical analyses are performed by different personnel.

The first step can be achieved by subtracting the mean of the measurements in one batch from all measurements in that batch, i.e mean adjustment or one-way ANOVA adjustment as implemented in the method \texttt{pamr.batchadjust} from the \texttt{pamr} package in R. When the batch--group design is balanced, mean-adjustment will remove most, but not necessarily all, variance attributed to batch and leave the between group variance, thus increasing the statistical power. However, when the batch--group design is unbalanced, batch differences will in part be influences by group differences, and thus batch correction will reduce group differences and thereby reduce the statistical power. In very uneven group--batch designs with multiple groups, spurious group differences may even be induced in this way. Figure~\ref{fig:boxplots} illustrates both these effects.

To mitigate the above problems, one may simultaneously estimate batch effects and group differences, e.g. using a two-way ANOVA, and only remove the batch differences from the data. Effectively, this means group differences are estimated based on within batch comparison, and applied to the batch adjusted data. In a balanced group--batch design, group differences and batch effects are independent, and this approach becomes identical to the above described zero-centering per batch. If the group--batch design is heavily unbalanced, estimation of group differences and batch effects are interdependent. Unfortunately, when group differences estimated from an unbalanced group--batch design are applied to the entire data set, if this batch adjusted data set is later analysed for group differences, the confidence will be exaggerated as the estimation of the group differences in the unbalanced design is less accurate than in the corresponding balanced design. Figure~\ref{fig:boxplots} illustrates how statistical uncertainties are deflated by this batch adjustment method by comparing them to the uncertainties from the original ANOVA.

The ComBat method described in \citet{Johnson2007}, and included in the sva package (\citealp{Leek2012}), can use either of the two above described approaches to estimate batch differences, but uses an empirical Bayes approach to avoid over-correction for batch effects for small batches. It has popularised the two-way ANOVA procedure for retaining group differences when adjusting for batch effects, and the inclusion of group difference as a covariate when removing batch effects has been recommended in the sva tutorial and user fora. Based on actual use of ComBat by the authors and others, ourselves included, we suspect thus adjusted data are commonly treated as "batch effect free" in subsequent analyses. And as a consequence, confidence in group effects has been overestimated and false results reported.

\NB{Kort oppsummere behovet for batch-korrigering: generelt og i microarraydata.}

\NB{Kort beskrive metoder som gjoer dette (batch-sentrering, ANOVA, ComBat).}

\NB{Forklare tilsiktet effekt disse paa batch+gruppe-forskjeller (balansert versus ikke-balansert).}

\NB{Forklare kort hva faktisk effekt er: ANOVA gir redusert gruppe-forskjeller, mens for ComBat overestimeres den statistiske styrken/sikkerheten av forskjellen.}

@@

We will base our discussion of data with batch effects on a simple model:
\begin{equation}
Y_{ijr}=\alpha+\beta_j+\gamma_i+\epsilon_{ijr}
\end{equation}
where $i=1,\ldots,m$ are the different batches, $j=1,\ldots,M$ are different study groups that we wish to compare, and $r=1,\ldots,n_{ij}$ are the different samples within batch $i$ and group $j$.

@@

For simplicity, we consider the case with a single gene and constant scale, i.e. $\delta_{ig}=1$. We will discuss the effect of empirical Bayes estimation of $\gamma_{ig}$ later, but out main argument is more easily made in the simpler context.

@@

The main ambition of batch effect adjustments is to be able to remove batch differences in such a way that downstream analyses of the adjusted data may be done without further batch adjustments.

The most common method for removing batch effects is to zero-centre each batch:
\begin{equation}
\Delta\widetilde Y_{ijr}=Y_{ijr}-\bar{Y}_{i}
\quad\text{where}\quad
\bar{Y}_i=\frac{\sum_{j=1}^M\sum_{r=1}^{n_{ij}} Y_{ijr}}{\sum_{j=1}^M n_{ij}}.
\end{equation}
An alternative is to centre each batch to the common average by adding the average value $\bar{Y}$ across the entire data set: i.e. $\widetilde Y_{ijr}=\Delta\widetilde Y_{ijr}+\bar{Y}$. When comparing groups, the common value $\bar{Y}$ has no effect, and so this is equilvalent to zero-centring each batch. However, if the different groups are unevenly represented in the different batches, the batch average $\bar{Y}_i$ will tend to capture group differences as well as batch effects. Thus, batch centering may reduce group differences, and thus reduce the power of downstream analyses.

Removing batch effects while retaining group differences can be done through an ANOVA analysis in which the group effects, $\beta_j$, and the batch effects, $\gamma_i$, are estimated simultaneously. Batch adjusted values may then be obtained by subtracting the estimated batch effects, $\hat\beta_j$:
\begin{equation}
\widetilde Y_{ijr}=Y_{ijr}-\hat\beta_j=\alpha+(\beta_j-\hat\beta_j)+\gamma_i+\epsilon_{ijr}.
\end{equation}
This will yield batch adjusted values where any systematic bias induced by the batch differences has been removed, while the group differences are retained.

Unfortunately, the estimation error $\hat\beta_j-\beta_j$ affects all values within the same batch in the same manner, inducing a dependency between the values. If the study groups are evenly represented in all batches, this will not influence estimated group differences as all groups are equally affected. However, if groups are unevenly represented, this induced dependency can have severe impact on downstream analyses.

@@

The undesired consequences of preserving group effects when correcting for batch effect is readily illustrated with a sanity check using random numbers . The documentation accompanying the sva library has a runnable example demonstrating how to adjust a data set with ComBat followed by a F-test. Swapping the real data with random numbers from a normal distribution (mean=0, sd=1), but otherwise following the instructions, will generate the p-value distribution shown in Figure 1. The skewed distribution is a indication that this approach may have a unintentional adverse effect. \NOTE{Jeg har ikke beskrevet permutatsjon checken fordi jeg er usikker paa om vi skal bruke den.}

%% FIGURE 1: Original position

@@

As the amount of false positive results when trying to retain group differences depends on the batch/group balance, we will show two examples with varying degree of unbalancedness.

1) In the first experiment(\citealp{Towfic2014}), cells were treated with glatiramer acetate (a medicine for multiple sclerosis) or a generic and mRNA was measured using microarrays alongside control samples. A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip) was observed and adjusted for with ComBat, whereafter the data was tested for differetially expressed genes, yielding a list of 1000 genes (Table S5, \citealp{Towfic2014}). Unfortunately the batch/treatment design was unbalanced with several batches having only one of the main treatments of interest. When we re-analyzed their data without using ComBat, but instead blocked for batch effect in limma, only 9 genes were found ($\text{FDR}<0.05$). Additional sanity checks with random numbers or perumted labels were also carried out and the distribution of p-values for differents settings are shown in Fig. 2.  Our conclusion is that most of the genes reported as differtially expresssed in (\citealp{Towfic2014}) are false positives. This example is a sort of "worst case" scenario for applying ComBat, since it both has a very unbalanced batch/group design and a a priori assumpotion of no difference. The R-code for our analysis and a more extensive report can be downloaded from github (\href{https://github.com/vegardny/combat_tests.git}).

%% FIGURE 2: Original position

2) The second example is taken from the supporting information for the original ComBat article (\citealp{Johnson2007}) where it is denoted "Data Set 2". Cells inhibited for the expression of the TAL1 gene were compared to controls on a microarray platform. The experiment was conducted on three different time points (used as batches) with a total of 30 samples and a fairly balanced batch/treatment set up (6-2, 3-4 and 9-6). ComBat was applied followed by a t-test in order to identify differentially expressed genes. First, we reproduced their analysis including the adjustment by ComBat, but using limma instead of the t-test, resulting in 1003 probes ($q<0.05$).  Then, we analysed their data without batch adjustment in ComBat, but blocking for batch in limma, resulting in 377 probes ($q<0.05$). In addition, the two sanity checks outlined above were performed. In contrast to the above results obtained for(\citealp{Towfic2014}), the P-value distributions for the alternative analyses (Fig.3 a,b) does not indicate a huge difference. Nevertheless, we belive that  P-values are deflated for the ComBat adjusted analysis. The R-code for our analysis and a more extensive report can be downloaded from github (\href{https://github.com/vegardny/combat_tests.git}).

%% FIGURE 3: Original position

%% FIGURE 4: Original position

3) \NOTE{Her kan vi fylle paa med flere eksempler etter samme mal hvis det trengs. Feks en fersk artikkel fra Nature Genetic (\citealp{Yoo2014}). Der har de kombinert sekvenseringsdata  (deres data) fra 9 lymfom pasienter med et (eksternt) microarray data set som har 6 lymfom proever og 5 kontroller, dvs kontroll gruppen er helt fravaerende hos dem. Dvs veldig ubalansert (9/0, 6/5), MEN skille mellom phenotypene er saa stor at de sikkert hadde funnet omtrent de samme genene signifikante med aa bare bruke de balanserte dataene (men det var jo ikke deres.) Dette er ikke en viktig del av paperet, men det er veldig feil selvom konsekvensense ikke blir store. Dog gir det et falskt inntrykk av at dataene deres ligner tidligere data, se heatmap i suppl.. Dessuten lager de GO analyser av dette som egentlig er 7 aar gamle andres data og ikke deres}

@@

@@

In the original ComBat article (\citealp{Johnson2007}) it is clear that the primary motivation behind ComBat was to employ an Emprical Bayes method for batch-effect removal. The feature of retaining group differences for unbalanced designs is optional and seems to be subordinate, only exemplified in the supplementary information. However, over the years this feature became more important judging from advice given by the author to users
(
[1](\href{https://groups.google.com/d/msg/combat-user-forum/eSVSKwGtuyE/ZIWV2juYmmAJ})
,[2](\href{https://groups.google.com/d/msg/combat-user-forum/Vkb9p7wekd4/h5Etie7FTVQJ})
,[3](\href{https://groups.google.com/d/msg/combat-user-forum/fpBTcgDjiR8/uo4QIZL4sZgJ})
,[4](\href{https://groups.google.com/d/msg/combat-user-forum/26FZlgU2LFQ/W6U_Lhh_64EJ})
). \NOTE{Aa linke direkte er kanskje litt ufint, alternativt kan det linkes til forumet uten konkrete innlegg.}  
And when ComBat was incorporated in the sva package (\citealp{Leek2012}), inclusion of group labels was made almost mandatory (an undocumented option of passing a NULL value exists). Thus, this problematic use has likely been common.

@@

Our knowledge of ComBat came through a typical use case when trying to salvage unbalanced data which had batch effects. Upon realizing that the confidence on our group differences were exaggerated, the literature was searched for a better understanding of correct use and potential overseen limitations of ComBat. But the authors of ComBat and the sva package recommended our usage (\citealp{Johnson2007}, \citealp{Leek2012}). In addition, other works looking into the problem of batch effects were mostly recommending ComBat without much concern (\citealp{Kupfer2012}, \citealp{Kitchen2011}). A brief inquiry into  some of the articles citing ComBat (574 Google Scholar) revealed few problems, and their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences (\CITE{24391845}, \CITE{18414638}, \CITE{21731603}, \CITE{23630272}, \CITE{24584070}). A further indication of their carefree use of this potentially devastating procedure was the frequent neglect to state the program parameters, i.e batch labels (\CITE{18414638}, \CITE{21731603}) or group labels (\CITE{24391845}, \CITE{18414638}, \CITE{21731603}, \CITE{23630272}, \CITE{23482648}, \CITE{24584070}). Often no effort was done in order to substantiate the existence of batch effects in their data, except for stating the presence of batches (\CITE{24391845}?, \CITE{18414638}, \CITE{21731603}, \CITE{23630272}). In one instance ComBat was even applied on data were effects due to batch were investigated, but not found \NOTE{husker ikke ref}. The incorporation of the method into analysis pipelines (\CITE{16642009}, TCGA) and other packages (\CITE{23452776}, \CITE{21937664}) could make its usage even more trivial and parameters setting harder to perceive.
Taken together we fear that many published results from data adjusted by ComBat are completely or partially false. And knowing that scientists don't give up where one analysis fails, a method that almost ensures a result given a sufficiently unbalanced design will continue to be used.

@@

We have shown that adjusting for batch effects while preserve the group difference may lead to varying degree of false results. Knowing this, to what degree can an investigator trust a result from a work applying such a method? Essentially, when the batch/group configuration is balanced, or group difference is ignored  (i.e. no group labels not given as parameters to ComBat), problems related to preserving group differences will not occur. For other cases, a re-analysis without using this approach is the most rigorous path. However, this thoroughness is not feasible if the downstream  analysis can not adjust for batch effects by it self. To reach a reliable result, batch effects need to be handled in some way or another. To make matters worse, a re-analysis relies on the availability of the raw data and a description of processing and analysis steps taken in the original work. Even when this is available, the necessary bioinformatic skills and work hours could still be in short supply. For such situations, a superficial assessment can be performed, taking special note of batches were groups of interest are near missing and how likely a group difference is. In essence asking if the balanced parts (effective sample size?) of the data has enough power to detect the presumed effects and if this is the case (\citealp{Johnson2007}, \CITE{24584070}), treat the results more like an ordered list with the most likely true positives on top while de-emphasizing the somewhat deflated p-values . In contrast, if biological knowledge suggest that a group effect is unlikely  (\CITE{24391845}, \CITE{18414638}?, \CITE{21731603}, \CITE{23630272}), an intermediate lack of batch/group balance  could lead to a mostly false result.

@@

\section*{Reproducible research}
\NOTE{vi boer klare aa tilfredstille kravene i http://biostatistics.oxfordjournals.org/content/10/3/405.full}

@@

\begin{figure}[!p]
\centering\includegraphics[width=13cm]{Fig/boxplots.png}
\caption{Simulated data from four study groups was generated where groups 1 and 2 have lower means than groups 3 and 4. These were placed in three different batches with batch effect added. Values and boxes showing mean and two standard errors of the mean are displayed for data without batch effects, after adding batch effects, after batch centering, and after ANOVA based batch centering. The last frame shows the least squares estimates of the group means from a two-way ANOVA analysis with 2 standard errors. This case, design and effects, was selected to illustrate the spurious effects that may arise.}
\label{fig:boxplots}
\end{figure}

\begin{figure}[!p]
\centering\includegraphics[width=13cm]{Fig/pvaluesleeksva.png}
\caption{
Two sanity checks were the recommended use of ComBat fails. Adapted from the user guide in the sva package. 
a) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a F-test.
b) 10 runs of real data with the "cancer" labels permuted within batches, followed by ComBat adjustment and a F-test. All permutations produces a skewed p-value distribution.
}
\label{fig:p-sva}
\end{figure}

>>>>

%% NOTES TO SELF:
% Cite using:
%    \citet{...}  in text
%    \citealp{...}   within parantheses
%
% Temporary notifications and reminders:
\usepackage{color}
\definecolor{NBcol}{rgb}{0.7,0,1}
\definecolor{CITEcol}{rgb}{1,0,0.7}
\definecolor{COMMENTcol}{rgb}{0.7,0.2,0}
\newcommand\COMMENT[2]{\textcolor{COMMENTcol}{#1}\NOTE{#2}}
\newcommand\NB[1]{\textcolor{NBcol}{\textit{#1}}}
\newcommand\NOTE[1]{\NB{[#1]}}
\newcommand\CITE[1]{\textcolor{CITEcol}{[#1]}}

@@

\footnotetext{\color{blue}Manuscript version \today.}

@@

% List of authors, with corresponding author marked by asterisk
\author{VEGARD NYGAARD, EINAR ANDREAS RØDLAND, EIVIND HOVIG$^\ast$\\[4pt]
% Author addresses
\textit{Department of Tumor Biology,
Institute for Cancer Research,
Oslo University Hospital HF - Radiumhospitalet,
Montebello,
0310 Oslo,
Norway}
\\[2pt]
% E-mail address for correspondence
{ehovig@ifi.uio.no}}

@@

The standard way to handle an extraneous variable is to include it in the statistical model employed in the inquiry. However, many analysis tools for high throughput data do not cater for this option, and when available it could still be outside the competence of the investigator. Therefore, an alternative two step procedure has emerged. First the batch effects are estimated and removed, creating a ``batch effect free'' data set. In the next step, the statistical analyses are performed on the adjusted data without further consideration of batch effects. This appealing compartmentalization is also convenient for practical purposes, for example when data-processing and statistical analyses are performed by different personnel. Unfortunately, as we demonstrate in this paper, when the batch--group design is unbalanced, this approach may be unreliable.

A simple removal of batch effects can be achieved by subtracting the mean of the measurements in one batch from all measurements in that batch, i.e mean adjustment or one-way ANOVA adjustment as implemented in the method \texttt{pamr.batchadjust} from the \texttt{pamr} package in R. When the batch--group design is balanced, mean-adjustment will remove most, but not necessarily all, variance attributed to batch and leave the between group variance, thus increasing the statistical power. However, when the batch--group design is unbalanced, batch differences will in part be influenced by group differences, and thus batch correction will reduce group differences and thereby reduce the statistical power. In very uneven group--batch designs with multiple groups, spurious group differences may even be induced in this way. Figure~\ref{fig:boxplots} illustrates both these effects.

To mitigate the above problems, one may simultaneously estimate batch effects and group differences, e.g. using a two-way ANOVA, and only remove the batch differences from the data. Effectively, group differences are estimated based on within batch comparisons, and applied to the batch adjusted data. In a balanced group--batch design, group differences and batch effects are independent, and this approach becomes identical to the above described zero-centering per batch. If the group--batch design is heavily unbalanced, estimation of group differences and batch effects are interdependent. However, in applying the estimated group differences across the entire data set, the uncertainties of these estimates are ignored. If the group--batch design is unbalanced and batch compositions ignored after batch adjustments have been made, later analyses will systematically underestimate the statistical uncertainties and exaggerate the confidence of group differences. Figure~\ref{fig:boxplots} illustrates how statistical uncertainties are deflated by this batch adjustment method by comparing them to the uncertainties from the original ANOVA.

The ComBat method described in \citet{Johnson2007}, and included in the sva package (\citealp{Leek2012}), can use either of the two above described approaches to estimate batch differences, but uses an empirical Bayes approach to avoid over-correction for batch effects for small batches. It has thus improved and popularised the two-way ANOVA procedure for retaining group differences when adjusting for batch effects, and the inclusion of group difference as a covariate when removing batch effects has been recommended  in the sva tutorial and user fora. Based on actual use of ComBat by the authors and others, ourselves included, we suspect thus adjusted data are commonly treated as ``batch effect free'' in subsequent analyses. And as a consequence, confidence in group effects has been overestimated and false results reported.

Statisticians would most likely take extra precautions should the sample or batch sizes be very small. However, the effects of batch adjustment using two-way ANOVA or ComBat on unbalanced data sets remain even as sample and batch sizes increase. For example, group comparisons using one-way ANOVA on the batch adjusted data will essentially result in $F$ statistics that are inflated by a fixed factor which depends on the unevenness of the design rather than the size of the sample or batches. The effect of this may be further exacerbated by running these analyses a large number of times, e.g. on thousands of genes, and use false discovery rate to determine significant cases: an approach that is particularly sensitive to inflated false positive rates.

@@

We will base our discussion on a simple model for data with batch effects:
\begin{equation}
Y_{ijr}=\alpha+\beta_j+\gamma_i+\epsilon_{ijr}
\end{equation}
where $i=1,\ldots,m$ are the different batches, $j=1,\ldots,M$ are different study groups that we wish to compare, and $r=1,\ldots,n_{ij}$ are the different samples within batch $i$ and group $j$.

@@

For simplicity, we consider the case with a single gene and constant scale, i.e. $\delta_{ig}=1$. We will discuss the effect of empirical Bayes estimation of $\gamma_{ig}$ later, but our main argument is more easily made in the simpler context. Empirical Bayes may reduce the amount of batch adjustment in cases where batch effects are small or cannot be accurately estimated, e.g. for small batches. However, in cases with large batches or substantial batch effects, it should differ little from the two-way ANOVA approach.

@@

A common ambition of batch effect adjustments is to remove batch differences in such a way that downstream analyses of the adjusted data may be done without further corrections for batches.

The most common method for removing batch effects is to zero-centre each batch:
\begin{equation}
\Delta\widetilde Y_{ijr}=Y_{ijr}-\bar{Y}_{i}
\quad\text{where}\quad
\bar{Y}_i=\frac{\sum_{j=1}^M\sum_{r=1}^{n_{ij}} Y_{ijr}}{\sum_{j=1}^M n_{ij}}.
\end{equation}
An alternative is to centre each batch to the common average by adding the average value $\bar{Y}$ across the entire data set: i.e. $\widetilde Y_{ijr}=\Delta\widetilde Y_{ijr}+\bar{Y}$. When comparing groups, the common value $\bar{Y}$ has no effect, and so this is equilvalent to zero-centring each batch. If the groups are unevenly represented in the different batches, the batch average $\bar{Y}_i$ will tend to capture group differences as well as batch effects:
\begin{equation}
\Delta\widetilde{Y}_{ijr}=\beta_j-\bar{\beta}_i+\epsilon_{ijr}-\bar{\epsilon}_i
\quad\text{where}\quad
\bar{\beta}_i=\frac{\sum_{j=1}^M n_{ij}\beta_j}{\sum_{j=1}^M n_{ij}},\;
\bar{\epsilon}_i=\frac{\sum_{j=1}^M\sum_{r=1}^{n_{ij}} \epsilon_{ijr}}{\sum_{j=1}^M n_{ij}}.
\end{equation}
Thus, batch centering will tend to reduce group differences in an unbalanced design, and thus reduce the power of downstream analyses. By reducing the differences between some groups, i.e. those found together in the same batches, it may also induce false differences between other groups.

Removing batch effects while retaining group differences can be done through a two-way ANOVA in which group effects, $\beta_j$, and batch effects, $\gamma_i$, are estimated simultaneously. Batch adjusted values may then be obtained by subtracting the estimated batch effects, $\hat\gamma_i$:
\begin{equation}
\widetilde Y_{ijr}=Y_{ijr}-\hat\beta_j=\alpha+\beta_j+(\gamma_i-\hat\gamma_i)+\epsilon_{ijr}.
\end{equation}
This will yield batch adjusted values where any systematic bias induced by the batch differences has been removed, while the group differences are retained.

The estimation error $\hat\gamma_i-\gamma_i$ affects all values within the same batch in the same manner. Thus, while the aim is to remove spurious dependencies within batches, it may also induce new dependencies. In a balanced group--batch design, this has the same effect on all groups and thus does not influence group comparisons. In an unbalanced design, however, the estimated batch effects $\hat\gamma_i$ will correlate with the estimated group effects $\hat\beta_j$. \COMMENT{Assuming the batch adjusted values are without further batch differences corresponds to assuming that $\gamma_i=\hat\gamma_i$, and ignoring the estimation uncertainty and its impact on the correlated $\hat\beta_j$. It results in over-confidence in the accuracy of the group differences, which can have severe impact on downstream analyses.}{Denne forklaringen må jeg gjøre mye klarere.}

@@

The undesired consequences of preserving group effects when correcting for batch effect is readily illustrated with a sanity check using random numbers . The documentation accompanying the sva library has a runnable example demonstrating how to adjust a data set with ComBat followed by a F-test. Swapping the real data with random numbers from a normal distribution (mean=0, sd=1), but otherwise following the instructions, will generate the p-value distribution shown in Figure~\ref{fig:p-sva}. \NOTE{Perhaps show Q-Q plot of $F$-statistics.} The skewed distribution is a indication that this approach may have a unintentional adverse effect. \NOTE{Tror vi dropper permutasjonstesten.}

If the number of samples is increased, the empirical Bayes estimates used by ComBat will shrink the batch effect estimates and thus moderate the batch adjustments. However, if random data with different means in each batch, but no difference betwene groups, are drawn, the problem remains. \NOTE{Bør dette komme her eller senere? Evt. understøttes?}

%% FIGURE 1=fig:p-sva: Original position

@@

As the amount of false positive results when trying to retain group differences depends on the batch--group balance, we will show two examples with varying degree of unbalancedness.

\subsubsection{Experiment 1}

In the first experiment(\citealp{Towfic2014}), cells were treated with glatiramer acetate (a medicine for multiple sclerosis) or a generic and mRNA was measured using microarrays alongside control samples. A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip) was observed and adjusted for with ComBat, whereafter the data was tested for differentially expressed genes, yielding a list of 1000 genes (Table S5, \citealp{Towfic2014}). Unfortunately the batch--treatment design was unbalanced with several batches having only one of the main treatments of interest. When we re-analyzed their data without using ComBat, but instead blocked for batch effect in limma, only 9 genes were found ($\text{FDR}<0.05$). \NOTE{Also run analysis using a mixed random effects model which assumes random batch effects much like in the empirical Bayes model.} The above outlined sanity check with random numbers was also carried out. The distribution of p-values for different settings are shown in Figure~\ref{fig:p-towfic}.  Our conclusion is that most of the genes reported as differentially expressed in (\citealp{Towfic2014}) are false positives. This example is a sort of ``worst case'' scenario for applying ComBat, since it both has a very unbalanced batch--group design and a a priori assumption of no difference. The R-code for our analysis and a more extensive report can be downloaded from \href{https://github.com/vegardny/combat_tests.git}{Github} \NOTE{Update reference}.

%% FIGURE 2=fig:p-towfic: Original position

\subsubsection{Experiment 2}

The second example is taken from the supporting information for the original ComBat article (\citealp{Johnson2007}) where it is denoted ``Data set 2''. Cells inhibited for the expression of the TAL1 gene were compared to controls on a microarray platform. The experiment was conducted on three different time points (used as batches) with a total of 30 samples and a fairly balanced batch--treatment set up (6-2, 3-4 and 9-6). ComBat was applied followed by a t-test in order to identify differentially expressed genes. First, we reproduced their analysis including the adjustment by ComBat, but using limma instead of the t-test, resulting in 1003 probes ($q<0.05$).  Then, we analysed their data without batch adjustment in ComBat, but blocking for batch in limma, resulting in 377 probes ($q<0.05$). In addition, the sanity checks outlined above were performed. The distribution of P-values for different settings are shown in Figure~\ref{fig:p-values}b. In contrast to the results obtained for\citealp{Towfic2014} (Figure~\ref{fig:p-values}a), the P-value distributions for the alternative analysis does not indicate a huge difference. Nevertheless, we belive that  P-values are deflated for the ComBat adjusted analysis. The R-code for our analysis can be downloaded from \href{https://github.com/vegardny/combat_tests.git}{Github} \NOTE{Update reference}.

%% FIGURE 3=fig:p-johnson: Original position

%% FIGURE 4=fig:pdist-johnson: Original position

\subsubsection{More?}

\NOTE{Her kan vi fylle paa med flere eksempler etter samme mal hvis det trengs. Feks en fersk artikkel fra Nature Genetic (\citealp{Yoo2014}). Der har de kombinert sekvenseringsdata  (deres data) fra 9 lymfom pasienter med et (eksternt) microarray data set som har 6 lymfom proever og 5 kontroller, dvs kontroll gruppen er helt fravaerende hos dem. Dvs veldig ubalansert (9/0, 6/5), MEN skille mellom phenotypene er saa stor at de sikkert hadde funnet omtrent de samme genene signifikante med aa bare bruke de balanserte dataene (men det var jo ikke deres.) Dette er ikke en viktig del av paperet, men det er veldig feil selvom konsekvensense ikke blir store. Dog gir det et falskt inntrykk av at dataene deres ligner tidligere data, se heatmap i suppl.. Dessuten lager de GO analyser av dette som egentlig er 7 aar gamle andres data og ikke deres}

@@

\NB{Systematic underestimation of $P$-values combined with FDR is a particularly bad combination as FDR lends strength across tests.}

@@

In the original ComBat article (\citealp{Johnson2007}) it is clear that the primary motivation behind ComBat was to employ an Emprical Bayes method for batch-effect removal to allow better handling of small batches. The feature of retaining group differences for unbalanced designs is optional and seems to be subordinate, only exemplified in the supplementary information. However, over the years this feature became more important judging from advice given on user fora.  
When ComBat was later incorporated in the sva package (\citealp{Leek2012}), \COMMENT{batch adjustment without covariates was no longer a default as specification of covariates became a requirement, although specification of a null model was still possible although not clearly documented in the sva documentation}{Is this correct?}. Thus, we expect this usage to have grown more common.

@@

Our knowledge of the problem discussed in this article came through a typical use case when trying to adjust for batch effects in an unbalanced data set using ComBat. Upon realizing that the confidence on our group differences were exaggerated, the literature was searched for a better understanding of correct use and potential overseen limitations of ComBat. But the authors of ComBat and the sva package recommended our usage (\citealp{Johnson2007}, \citealp{Leek2012}). In addition, other works looking into the problem of batch effects were mostly recommending ComBat without much concern (\citealp{Kupfer2012}, \citealp{Kitchen2011}). A brief inquiry into  some of the articles citing ComBat (574 Google Scholar) revealed few problems, and their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences. A further indication of their carefree use of this potentially devastating procedure was the frequent omittance of program parameters that were used, i.e batch labels or whether group labels were supplied as covariates. Often no effort was done in order to substantiate the existence of batch effects in their data, except for stating the presence of batches. The incorporation of the method into analysis pipelines (\CITE{16642009}, TCGA) and other packages (\CITE{23452776}, \CITE{21937664}) could make its usage even more trivial and parameters setting harder to perceive. Taken together we fear that many published results from data adjusted by ComBat are completely or partially false. We hope that our warning will enlighten the community and  reduce this unfortunate combination of methods.

@@

For an investigator facing an unbalanced data set with batch effects, our advice is to adjust for batch inside the statistical test and avoid the two step procedure outlined above. If this is not possible, then applying the method described in this article should only be done with great caution and downstream confidence estimates must be treated with suspicion.

Knowing that adjusting for batch effects while preserving the group difference may lead to varying degree of false results, to what extent can an investigator trust a result from a work applying such a method? Essentially, when the batch--group configuration is balanced, or group difference is ignored  (i.e. group labels not given as parameters to ComBat), problems related to preserving group differences will not occur. For other cases, a re-analysis without using this approach is the most rigorous path. However, this thoroughness is not feasible if the downstream  analysis can not adjust for batch effects by it self. To reach a reliable result, batch effects need to be handled in some way or another. To make matters worse, a re-analysis relies on the availability of the raw data and a description of processing and analysis steps taken in the original work. Even when this is available, the necessary bioinformatic skills and work hours could still be in short supply. For such situations, a superficial assessment can be performed, taking special note of batches were groups of interest are near missing and how likely a group difference is. In essence asking if the balanced parts (effective sample size?) of the data has enough power to detect the presumed effects and if this is the case (\citealp{Johnson2007}, \CITE{24584070}), treat the results more like an ordered list with the most likely true positives on top while de-emphasizing the somewhat deflated p-values . In contrast, if biological knowledge suggest that a group effect is unlikely  (\CITE{24391845}, \CITE{18414638}?, \CITE{21731603}, \CITE{23630272}), an intermediate lack of batch-group balance  could lead to a mostly false result.

@@

\subsection{Reproducible research}

The data and scripts used to generate the results in this work, i.e figures, are made available at https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git. Additional analyses were also performed, but considered excessive for the current  article. However, some readers might find this very interesting and we provide it as an extra resource at https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-reports.git

\NOTE{vi boer klare aa tilfredstille kravene i http://biostatistics.oxfordjournals.org/content/10/3/405.full}
\NOTE{For aa tilfredstille kravene skal data og script vare available. Jeg mener vi klarer dette. paa githuben er ikke data fra towfic, men det blir lastet av scriptet fra GEO. Ikke beskrevet her men i scriptet. Videre i forklaringen til biostatistics staar det at script og data skal bli lagt paa deres sider, som vi kan hvis de vil}

@@

\begin{figure}[!p]
\centering\includegraphics[width=13cm]{Fig/boxplots.pdf}
%\centering\includegraphics[width=13cm]{Fig/boxplots.png}
\caption{Simulated data from four study groups was generated where groups 1 and 2 have lower means than groups 3 and 4. These were placed in three different batches with batch effect added. Values and boxes showing mean and two standard errors of the mean are displayed for data without batch effects, after adding batch effects, after batch centering, and after ANOVA based batch centering. The last frame shows the least squares estimates of the group means from a two-way ANOVA analysis with 2 standard errors. This case, design and effects, was selected to illustrate the spurious effects that may arise.}
\label{fig:boxplots}
\end{figure}

\begin{figure}[!p]
\begin{center}$
\begin{array}{cc}
\includegraphics[width=6.5cm]{Fig/leekrandomdatapvalues.pdf} &
\includegraphics[width=6.5cm]{Fig/leekqqplot.png}
\end{array}$
\end{center}
\caption
{
A sanity checks where the recommended use of ComBat fails. Adapted from the user guide in the sva package. Real data are substituted with random numbers from a normal distribution, but the batch-group design is retained. ComBat is applied followed by a F-test.
a) P-value distribution
b) QQ plot of the F-statistics
}
\end{figure}

\begin{figure}[!p]
\begin{center}$
\begin{array}{cc}
\includegraphics[width=6.5cm]{Fig/towficpvalues.pdf} &
\includegraphics[width=6.5cm]{Fig/johnson2pvalues.pdf}
\end{array}$
\end{center}
\caption
{
Three analyses of two works where batch effects were adjusted for with ComBat. First as described in the original works with ComBat on real data. Secondly, with ComBat but with random numbers instead of real data. Thirdly without ComBat and instead blocking batch with limma on real data.
a) Re-analysis of  \citealp{Towfic2014}, glatiramer acetate vs. generic 
b) Re-analysis of "Data set 2" \citealp{Johnson2007},  TAL1 inhibition vs. control
}
\end{figure}
